Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-08-02 paper/ifo-complexAudio.tex
proof read

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing (folding of frequencies greater than half the sampling rate) into the detected range. We include an anti-aliasing Sallen-Key filter~\cite{sallen_key_filter} tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$ before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was re-purposed grill cloth from a commercial speaker.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

The raw output of the optical microphone (with anti-aliasing) is noisy with a loud bass continuous hum. 
This can be explained by looking at the power spectral density (PSD) of the background noise (i.e., with the speaker switched off), shown in Fig.~\ref{fig:psd_noise}. 
The spectrum is dominated by AC mains noise with power from the fundamental $50\,{\rm Hz}$ Australian mains power grid signal up to and beyond $8$th harmonic. 
The mains signal is also present, but far weaker, in the background spectrum taken with the photodiode in darkness, suggesting that ambient lighting has a large contribution. 
Besides lighting, other possible contributions to the mains signal include air conditioning and the photodiode circuit itself. 
The appearance of harmonics of the mains noise is likely due to non-linearity in the system (see also Ref.~\cite{feynman}). 
The spectrum in Fig.~\ref{fig:psd_noise} also has a broad feature at around $750\,{\rm Hz}$, the origin of which is yet to be determined.
Environmental noise reduction for gravitational-wave detectors is an active area of research (see the accompanying Supplementary Material for further information and resources on this topic).

@@

Firstly, we consider two signal processing techniques used in combination: the cascaded notch~\cite{10.5555/541204} and the Wiener filter~\cite{10.5555/151045}. 
A notch filter removes signals within a specific frequency range. 
We want to remove the $50\,{\rm Hz}$ mains noise and harmonics, therefore we use a cascaded notch filter where each notch is centered at each of the harmonics. 
The Wiener filter is an advanced statistical technique that makes use of statistical information from the speech data and noise. 
It amplifies parts of the signal with high signal-to-noise (SNR) ratio while suppressing parts with low SNR. 
The results of the combined cascaded notch and Wiener filter are shown in the third row in Fig.~\ref{fig:notchWienerLogMMSEResults}. 
Most of the mains noise is removed, however the recovered voice sounds muffled and is not understandable.

Secondly, we apply a speech enhancement technique. 
Ref.~\cite{SubjectiveComparison} compares $13$ speech enhancement methods, finding the log minimum mean-square error (logMMSE) estimator to be the best, qualitatively, at recovering speech (see also Ref.~\cite{Ephraim1984SpeechEU_logMMSE} for details on the logMMSE estimator). 
We use an existing implementation of the logMMSE from~\cite{logmmse}.  
The logMMSE estimator results are shown in the bottom panels in Fig.~\ref{fig:notchWienerLogMMSEResults}. 
We see significant attenuation of the mains harmonics and general smoothing of the spectrum. 
Most of the background noise is removed, however, the logMMSE does not significantly enhance the speech as the voice sounds muffled and indistinct.

>>>>

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing (folding of frequencies greater than half the sampling rate) into the detected range. We include an anti-aliasing Sallen-Key filter tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$ before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was re-purposed grill cloth from a commercial speaker.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

The raw output of the optical microphone (with anti-aliasing) is noisy with a loud bass continuous hum. 
This can be explained by looking at the power spectral density (PSD) of the background noise (i.e., with the speaker switched off), shown in Fig.~\ref{fig:psd_noise}. 
The spectrum is dominated by AC mains noise with power from the fundamental $50\,{\rm Hz}$ Australian mains power grid signal up to and beyond $8$th harmonic. 
The mains signal is also present, but far weaker, in the background spectrum taken with the photodiode in darkness, suggesting that ambient lighting has a large contribution. 
Besides lighting, other possible contributions to the mains signal include air conditioning and the photodiode circuit itself. 
The appearance of harmonics of the mains noise is likely due to non-linearity in the system. 
The spectrum in Fig.~\ref{fig:psd_noise} also has a broad feature at around $750\,{\rm Hz}$, the origin of which is yet to be determined.
Environmental noise reduction for gravitational-wave detectors is an active area of research (see the accompanying Supplementary Material for further information and resources on this topic).

@@

Firstly, we consider two signal processing techniques used in combination: the cascaded notch and the Wiener filter (see also the Supplementary Material). 
A notch filter removes signals within a specific frequency range. 
We want to remove the $50\,{\rm Hz}$ mains noise and harmonics, therefore we use a cascaded notch filter where each notch is centered at each of the harmonics. 
The Wiener filter is an advanced statistical technique that makes use of statistical information from the speech data and noise. 
It amplifies parts of the signal with high signal-to-noise (SNR) ratio while suppressing parts with low SNR. 
The results of the combined cascaded notch and Wiener filter are shown in the third row in Fig.~\ref{fig:notchWienerLogMMSEResults}. 
Most of the mains noise is removed, however the recovered voice sounds muffled and is not understandable.

Secondly, we apply a speech enhancement technique. 
Ref.~\cite{SubjectiveComparison} compares $13$ speech enhancement methods, finding the log minimum mean-square error (logMMSE) estimator to be the best, qualitatively, at recovering speech (see also the Supplementary Material). 
We use an existing implementation of the logMMSE from~\cite{logmmse}.  
The logMMSE estimator results are shown in the bottom panels in Fig.~\ref{fig:notchWienerLogMMSEResults}. 
We see significant attenuation of the mains harmonics and general smoothing of the spectrum. 
Most of the background noise is removed, however, the logMMSE does not significantly enhance the speech as the voice sounds muffled and indistinct.

### 2021-08-02 paper/ifo-wanderingTone.tex
proof read

In continuous wave searches, a physical model of the target informs how far the frequency of the signal can wander over time. 
This is called the transition probability matrix. 
For example, in LMXB searches, the transition matrix allows the signal frequency to (i) stay in the same frequency bin, (ii) move up a single frequency bin, or (iii) move down a frequency bin in each time step~\cite{ScoX1O2Viterbi:2019}. 
In supernova remnant searches, source frequency is expected to decrease over time, therefore the allowable transitions are to (i) remain in the same frequency bin, or (ii) move down one frequency bin (see the Supplementary Material for supernova remnant search references).

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/viterbiDiagramGrey.pdf}
\caption{\label{fig:viterbi}
Schematic diagram of the Viterbi algorithm. 
The circles represent elements in a time-frequency grid, labeled $t_0$ to $t_{N_t}$ (left to right) in time and $f_0$ to $f_{N_f}$ (bottom to top) in frequency. 
The shading of each circle represents the likelihood of a signal being present in each time-frequency element.
The dark to light gradient corresponds to low to high likelihood values in arbitrary units for this schematic. 
The objective is to find the most probable path of a signal through the grid from left to right.
All possible paths are shown by the lines. 
The black lines show the best path to each circle and the gray lines show rejected paths. 
Some routes through the grid result in dead-ends with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a).
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:pathprobability}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

Fig.~\ref{fig:viterbi} represents a spectrogram with $N_t$ time bins and $N_f$ frequency bins.
The circles represent the elements of the time and frequency grid, which are labelled as  
$t_i$ and $f_j$ respectively, where $i=0,1,2,...N_t$ and $j=0,1,2,...,N_f$. 
The shading of the circles represents the detection statistic $F(t_i,f_j)$ (the observable) at each grid point, where dark corresponds to lower values and light to higher values.

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heat-map shows the spectrogram of the observed signal (similar to that represented by the schematic in Fig.~\ref{fig:viterbi}). 
In this demonstration we use a signal that can easily be identified in the spectrogram, however, we expect real continuous-wave signals to have far lower signal-to-noise ratios. 
The overlaid pink dots in Fig.~\ref{fig:viterbi_overlay} show the injected signal and the white crosses show the recovered Viterbi path.
The recovered Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. We also compute the root-mean-square (RMS) fractional error $E_{\rm rms}$ along the path, which is defined as 
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}\,,
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths, respectively. 
For the result shown in Fig.~\ref{fig:viterbi_overlay}, we find that $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.
If we remove the first four time bins from the $E_{\rm rms}$ calculation, we find a slight improvement in recovery with $E_{\rm rms} = 0.044$.

This may be explained by two anomalies in the recovered path. Initially, the injected signal wanders by more than one frequency bin per time bin (i.e.\ faster than our choice of $A(f_k,f_m)$ allows), thus leading to a discrepancy between the injected and recovered paths for $t\lesssim 10\,{\rm s}$. One may be tempted to increase the allowed frequency wander in the algorithm, however, this leads to an overall decrease in the above statistics, as the algorithm is prone to jump briefly to nearby spots of noise. There is also an anomaly at $150\,{\rm s}$, which is likely due to a local disturbance, e.g. someone walking past the interferometer. 
As shown in Fig.~\ref{fig:viterbi_overlay}, the Viterbi algorithm can recover and continues to track the signal after the disturbance.

>>>>

In continuous-wave searches, a physical model of the target informs how far the frequency of the signal can wander over time. 
This is called the transition probability matrix. 
For example, in LMXB searches, the transition matrix allows the signal frequency to (i) stay in the same frequency bin, (ii) move up a single frequency bin, or (iii) move down a frequency bin in each time step~\cite{ScoX1O2Viterbi:2019}. 
In supernova remnant searches, source frequency is expected to decrease over time, therefore the allowable transitions are to (i) remain in the same frequency bin, or (ii) move down one frequency bin (see the Supplementary Material for LMXB and supernova remnant search references).

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/viterbiDiagramGrey.pdf}
\caption{\label{fig:viterbi}
Schematic diagram of the Viterbi algorithm. 
The circular nodes represent elements in a time-frequency grid, labeled $t_0$ to $t_{N_t}$ (left to right) in time and $f_0$ to $f_{N_f}$ (bottom to top) in frequency. 
The shading of each node represents the likelihood of a signal being present in each time-frequency element.
The dark to light gradient corresponds to low to high likelihood values in arbitrary units for this schematic. 
The objective is to find the most probable path of a signal through the grid from left to right.
All possible paths are shown by the lines. 
The black lines show the best path to each node and the gray lines show rejected paths. 
Some routes through the grid result in dead-ends with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a).
At $t=t_{N_t}$ the algorithm chooses the terminating frequency node which has the highest value given by Eqn.~\ref{eqn:pathprobability}, marked (b). 
The Viterbi path is the path leading to this node, highlighted in orange, from (b) to the start at (c). 
}
\end{figure}

@@

Fig.~\ref{fig:viterbi} represents a spectrogram with $N_t$ time bins and $N_f$ frequency bins.
The circlular nodes represent the elements of the time and frequency grid, which are labelled as  
$t_i$ and $f_j$, respectively where $i=0,1,2,...N_t$ and $j=0,1,2,...,N_f$. 
The shading of the nodes represents the detection statistic $F(t_i,f_j)$ (the observable) at each grid point, where dark corresponds to lower values and light to higher values.

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heat-map shows the spectrogram of the observed signal (similar to that represented by the schematic in Fig.~\ref{fig:viterbi}). 
In this demonstration we use a signal that can easily be identified in the spectrogram, however, we expect real continuous-wave signals to have far lower signal-to-noise ratios at each time step. 
The overlaid pink dots in Fig.~\ref{fig:viterbi_overlay} show the injected signal and the white crosses show the recovered Viterbi path.
The recovered Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. We also compute the root-mean-square (RMS) fractional error $E_{\rm rms}$ along the path, which is defined as 
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}\,,
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths, respectively. 
For the result shown in Fig.~\ref{fig:viterbi_overlay}, we find that $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.

This may be explained by two anomalies in the recovered path. Initially, the injected signal wanders by more than one frequency bin per time bin (i.e.\ faster than our choice of $A(f_k,f_m)$ allows), thus leading to a discrepancy between the injected and recovered paths for $t\lesssim 10\,{\rm s}$. 
If we remove the first four time bins from the $E_{\rm rms}$ calculation, we find a slight improvement in recovery with $E_{\rm rms} = 0.044$.
One may be tempted to increase the allowed frequency wander in the algorithm, however, this leads to an overall decrease in the above statistics, as the algorithm is prone to jump briefly to nearby spots of noise. There is also an anomaly at $150\,{\rm s}$, which is likely due to a local disturbance, e.g. someone walking past the interferometer. 
As shown in Fig.~\ref{fig:viterbi_overlay}, the Viterbi algorithm can recover and continues to track the signal after the disturbance.
